A comparison of design-based and model-based approaches for finite population spatial data.

### 2021-12-21 inst/manuscript/manuscript.Rmd
minor updates

---
title: A comparison of design-based and model-based approaches for finite population spatial data.
author:
  - name: Michael Dumelle
    affiliation: USEPA
    footnote: 1
  - name: Matt Higham
    affiliation: STLAW
  - name: Jay M. Ver Hoef
    affiliation: NOAA
  - name: Anthony R. Olsen
    affiliation: USEPA
  - name: Lisa Madsen
    affiliation: OSU
address:
  - code: USEPA
    address: United States Environmental Protection Agency, 200 SW 35th St, Corvallis, Oregon, 97333
  - code: STLAW
    address: Saint Lawrence University Department of Mathematics, Computer Science, and Statistics, 23 Romoda Drive, Canton, New York, 13617
  - code: NOAA
    address: Marine Mammal Laboratory, Alaska Fisheries Science Center, National Oceanic and Atmospheric Administration, Seattle, Washington, 98115
  - code: OSU
    address: Oregon State University Department of Statistics, 239 Weniger Hall, Corvallis, Oregon, 97331
footnote:
    code: 1
    text: "Corresponding Author: Michael Dumelle (Dumelle.Michael@epa.gov)"
abstract: |
  1.   The design-based and model-based approaches to frequentist statistical inference lie on fundamentally different foundations. In the design-based approach, inference depends on random sampling. In the model-based approach, inference depends on distributional assumptions. We compare the approaches for finite population spatial data.
  2. We provide relevant background for the design-based and model-based approaches and then study their performance using simulations and an analysis of real mercury concentration data. In the simulations, a variety of sample sizes, location layouts, dependence structures, and response types are considered. In the simualtions and real data analysis, the population mean is the parameter of interest and performance is measured using statistics like bias, squared error, and interval coverage.
  3. When studying the simulations and mercury concentration data, we found that regardless of the strength of spatial dependence in the data, sampling plans that incorporate spatial locations (spatially balanced samples) generally outperform sampling plans that ignore spatial locations (non-spatially balanced samples). We also found that model-based approaches tend to outperform design-based approaches, even when the data are skewed (and by consequence, the model-based distributional assumptions violated). The performance gap between the analysis approaches is small when spatially balanced samples are used but large when non-spatially balanced samples are used. This suggests that the sampling choice (whether to select a sample that is spatially balanced) is most important when performing design-based inference.
  4. There are many benefits and drawbacks to the design-based and model-based approaches for finite population spatial data that practitioners must consider when choosing between them. We provide relevant background contextualizing each approach and study their properties in a variety of scenarios, making recommendations for use based on the practitioner's goals.
journal: "Methods in Ecology and Evolution"
date: "`r Sys.Date()`"
bibliography: mybibfile.bib
linenumbers: true
numbersections: true
csl: elsevier-harvard.csl
preamble: >
  \usepackage{bm}
  \usepackage{bbm}
  \usepackage{color}
  \DeclareMathOperator{\var}{{var}}
  \DeclareMathOperator{\cov}{{cov}}
  \usepackage{caption}
  \usepackage{subcaption}
  \usepackage{setspace}
  \doublespacing
output: rticles::elsevier_article
---

>>>>

---
title: A comparison of design-based and model-based approaches for finite population spatial data.
author:
  - name: Michael Dumelle
    affiliation: USEPA
    footnote: 1
  - name: Matt Higham
    affiliation: STLAW
  - name: Jay M. Ver Hoef
    affiliation: NOAA
  - name: Anthony R. Olsen
    affiliation: USEPA
  - name: Lisa Madsen
    affiliation: OSU
address:
  - code: USEPA
    address: United States Environmental Protection Agency, 200 SW 35th St, Corvallis, Oregon, 97333
  - code: STLAW
    address: Saint Lawrence University Department of Mathematics, Computer Science, and Statistics, 23 Romoda Drive, Canton, New York, 13617
  - code: NOAA
    address: Marine Mammal Laboratory, Alaska Fisheries Science Center, National Oceanic and Atmospheric Administration, Seattle, Washington, 98115
  - code: OSU
    address: Oregon State University Department of Statistics, 239 Weniger Hall, Corvallis, Oregon, 97331
footnote:
    code: 1
    text: "Corresponding Author: Michael Dumelle (Dumelle.Michael@epa.gov)"
abstract: |
  1.   The design-based and model-based approaches to frequentist statistical inference lie on fundamentally different foundations. In the design-based approach, inference depends on random sampling. In the model-based approach, inference depends on distributional assumptions. We compare the approaches for finite population spatial data.
  2. We provide relevant background for the design-based and model-based approaches and then study their performance using simulations and an analysis of real mercury concentration data. In the simulations, a variety of sample sizes, location layouts, dependence structures, and response types are considered. In the simualtions and real data analysis, the population mean is the parameter of interest and performance is measured using statistics like bias, squared error, and interval coverage.
  3. When studying the simulations and mercury concentration data, we found that regardless of the strength of spatial dependence in the data, sampling plans that incorporate spatial locations (spatially balanced samples) generally outperform sampling plans that ignore spatial locations (non-spatially balanced samples). We also found that model-based approaches tend to outperform design-based approaches, even when the data are skewed (and by consequence, the model-based distributional assumptions violated). The performance gap between these approaches is small when spatially balanced samples are used but large when non-spatially balanced samples are used. This suggests that the sampling choice (whether to select a sample that is spatially balanced) is most important when performing design-based inference.
  4. There are many benefits and drawbacks to the design-based and model-based approaches for finite population spatial data that practitioners must consider when choosing between them. We provide relevant background contextualizing each approach and study their properties in a variety of scenarios, making recommendations for use based on the practitioner's goals.
journal: "Methods in Ecology and Evolution"
date: "`r Sys.Date()`"
bibliography: mybibfile.bib
linenumbers: true
numbersections: true
csl: elsevier-harvard.csl
preamble: >
  \usepackage{bm}
  \usepackage{bbm}
  \usepackage{color}
  \DeclareMathOperator{\var}{{var}}
  \DeclareMathOperator{\cov}{{cov}}
  \usepackage{caption}
  \usepackage{subcaption}
  \usepackage{setspace}
  \doublespacing
output: rticles::elsevier_article
---
